---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# vayr

<!-- badges: start -->
<!-- badges: end -->

The goal of `vayr` is to provide `ggplot2` extensions that foster "visualize as 
you randomize" principles, which should guide the visualization of experimental 
data. So far the package includes position adjustments to avoid over-plotting, 
facilitating the use of statistical models in "data-space." In other words,
these position adjustments help you organize "data-space" so that it can better 
contextualize your models. The principles underlying `vayr` are outlined in 
detail in "Visualize as You Randomize: Design-Based Statistical Graphs for 
Randomized Experiments." That paper can be accessed as a [PDF](https://alexandercoppock.com/coppock_2020.pdf).

## Installation

You can install the release version of `vayr` from CRAN, or you can install the 
development version from [GitHub](https://github.com/) using a package like 
`remotes`, `devtools `, or `pak`. It is important to note that `vayr` relies on 
`ggplot2`, `packcircles`, and `withr`, so you must have these installed as well.

``` r
# From CRAN
install.packages("vayr")

# From GitHub
# install.packages("pak")
pak::pak("acoppock/vayr")
```

## Contents

The current version of `vayr` contains a handful of `ggplot2` functions that you 
can apply as position adjustments to "point-like" geoms such as `geom_point` or 
`geom_text`:

  + `position_jitter_ellipse()` and `position_jitterdodge_ellipse()`
  + `position_sunflower()` and `position_sunflowerdodge()`
  + `position_circlepack()` and `position_circlepackdodge()`
  
These functions avoid over-plotting, so they are especially useful when you are 
plotting discrete rather than continuous data in a bivariate scenario. As a matter 
of demonstration, they are used below to visualize synthetic data, over-plotted
at the origin.

```{r, echo = FALSE}
suppressMessages(library(dplyr))
```

```{r setup}
library(dplyr)
library(estimatr)
library(ggplot2)
library(patchwork)
library(vayr)

dat <- data.frame(
  x = c(rep(0, 200)),
  y = c(rep(0, 200)),
  group = (rep(c("A", "B", "B", "B"), 50)),
  size = runif(200, 0, 1)
)
```

If position is the product of discrete variables alone, then over-plotting is of 
particular concern. To mitigate this concern, you can introduce variation with 
`position_jitter()`, which randomly samples points on a rectangle. This approach 
is effective but can be unattractive. The position adjustments in `vayr` are 
attempts to do better. 

```{r contents_0, dpi = 96, fig.width = 12, fig.height = 6, fig.alt = 'perfect over-plotting and position_jitter()'}
# perfectly over-plotted points
overplot <- ggplot(dat, aes(x = x, y = y)) +
  geom_point() +
  coord_equal(xlim = c(-1.1, 1.1), 
              ylim = c(-1.1, 1.1)) +
  theme_bw() +
  theme(axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle('"perfect over-plotting"')

# position_jitter()
plot <- ggplot(dat, aes(x = x, y = y)) + 
  geom_point(position = position_jitter(width = 0.5, 
                                        height = 0.5)) +
  coord_equal(xlim = c(-1.1, 1.1), 
              ylim = c(-1.1, 1.1)) +
  theme_bw() +
  theme(axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("position_jitter()")

overplot + plot
```

### Position Jitter Ellipse

This position adjustment adds elliptical random noise to perfectly over-plotted 
points, offering a pleasing way to visualize many points that represent the same 
position. The benefit of sampling on an ellipse (of a given `height` and `width`) 
rather than a rectangle is that the resulting dispersion retains the impression 
of a single point. The size of the ellipses stays constant, while their density 
varies depending on the amount of over-plotting.

```{r contents_1, dpi = 96, fig.width = 12, fig.height = 6, fig.alt = 'position_jitter_ellipse() and position_jitterdodge_ellipse()'}
# position_jitter_ellipse()
plot <- ggplot(dat, aes(x = x, y = y)) +
  geom_point(position = position_jitter_ellipse(width = 0.5, 
                                                height = 0.5)) +
  coord_equal(xlim = c(-1.1, 1.1), 
              ylim = c(-1.1, 1.1)) +
  theme_bw() +
  theme(axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("position_jitter_ellipse()")

# position_jitterdodge_ellipse()
dodged_plot <- ggplot(dat, aes(x = x, y = y, color = group)) +
  geom_point(position = position_jitterdodge_ellipse(dodge.width = 2, 
                                                     jitter.width = 0.5, 
                                                     jitter.height = 0.5)) +
  coord_equal(xlim = c(-1.1, 1.1), 
              ylim = c(-1.1, 1.1)) +
  theme_bw() +
  theme(legend.position = "none",
        axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("position_jitterdodge_ellipse()")
  
plot + dodged_plot
```

### Position Sunflower

This position adjustment arranges perfectly over-plotted points using a sunflower 
algorithm, which produces a pattern that resembles the seeds of a sunflower, 
working from the inside out in the order of the data. The parameters for this 
position adjustment (`density` and `aspect_ratio`) are novel but intuitive. 
The size of the flowers varies depending on the amount over over-plotting, but 
the density of the pattern remains constant. You can adjust the aspect ratio of 
the flowers, but it is generally recommended that the position adjustment be used 
along with `coord_equal()`, in which case the default aspect ratio of 1 yields 
perfectly circular flowers.

```{r contents_2A, dpi = 96, fig.width = 12, fig.height = 6, fig.alt = 'position_sunflower() and position_sunflowerdodge()'}
# position_sunflower()
plot <- ggplot(dat, aes(x = x, y = y)) +
  geom_point(position = position_sunflower(density = 1, 
                                           aspect_ratio = 1)) +
  coord_equal(xlim = c(-1.1, 1.1), 
              ylim = c(-1.1, 1.1)) +
  theme_bw() +
  theme(axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("position_sunflower()")
  
# position_sunflowerdodge()
dodged_plot <- ggplot(dat, aes(x = x, y = y, color = group)) +
  geom_point(position = position_sunflowerdodge(width = 2, 
                                                density = 1, 
                                                aspect_ratio = 1)) +
  coord_equal(xlim = c(-1.1, 1.1), 
              ylim = c(-1.1, 1.1)) +
  theme_bw() + 
  theme(legend.position = "none",
        axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("position_sunflowerdodge()")
  
plot + dodged_plot
```

As stated, you can adjust the density of the pattern. Density is not standardized
but rather depends on the ranges of the axes and the dimensions of the saved image. 
Loosely, you can think of it as the number of points within a certain Cartesian area.
Thus smaller ranges and/or larger dimensions require a greater density to produce 
the same visual effect.

```{r contents_2B, dpi = 96, fig.width = 12, fig.height = 4, fig.alt = 'density'}
# base plot
plot <- ggplot(dat, aes(x = x, y = y)) +
  coord_equal(xlim = c(-1.1, 1.1), 
              ylim = c(-1.1, 1.1)) +
  theme_bw() +
  theme(axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold"))

# low density
low <- plot + 
  geom_point(size = 0.25,
             position = position_sunflower(density = 0.5, 
                                           aspect_ratio = 1)) + 
  ggtitle("density = 0.5")

# medium density
medium <- plot + 
  geom_point(size = 0.25,
             position = position_sunflower(density = 1, 
                                           aspect_ratio = 1)) + 
  ggtitle("density = 1")

# high density
high <- plot + 
  geom_point(size = 0.25,
             position = position_sunflower(density = 2, 
                                           aspect_ratio = 1)) + 
  ggtitle("density = 2")

low + medium + high
```

If need be, you can also tweak the aspect ratio of the flowers. This is useful 
when you are using the position adjustment without `coord_equal()`. The flowers 
can be made wider or taller to compensate for the aspect ratio of the axes and/or 
the image. The aspect ratio of the flower should be set to the opposite of the 
aspect ratio for which it must compensate. For instance, consider a plot with an
x axis that ranges from 0 to 1, and a y axis that ranges from 0 to 2. Saving this
plot as a square image would effectively squish the y axis, resulting in wider 
flowers. Setting the aspect ratio of the flowers to 0.5 would offset this distortion, 
although further refinement would likely be warranted due to labels, legends, 
and the like.

```{r contents_2C, dpi = 96, fig.width = 12, fig.height = 4, fig.alt = 'aspect_ratio'}
# base plot
plot <- ggplot(dat, aes(x = x, y = y)) +
  coord_equal(xlim = c(-1.1, 1.1), 
              ylim = c(-1.1, 1.1)) +
  theme_bw() +
  theme(axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold"))

# wide aspect ratio 
wide <- plot + 
  geom_point(size = 0.25,
             position = position_sunflower(density = 1, 
                                           aspect_ratio = 2)) + 
  ggtitle("aspect_ratio = 2")

# normal aspect ratio
normal <- plot + 
  geom_point(size = 0.25,
             position = position_sunflower(density = 1, 
                                           aspect_ratio = 1)) + 
  ggtitle("aspect_ratio = 1")

# tall aspect ratio
tall <- plot + 
  geom_point(size = 0.25,
             position = position_sunflower(density = 1, 
                                           aspect_ratio = 0.5)) + 
  ggtitle("aspect_ratio = 0.5")

wide + normal + tall
```

### Position Circle Pack

This position adjustment uses circle packing algorithms from the `packcircles` 
package to arrange perfectly over-plotted points of varying sizes into an 
elliptical area. It also takes `density` and `aspect_ratio` as parameters, and 
they function as previously described. This position adjustment should not be
confused with `geom_circlepack()` from the `ggcirclepack` package, which can be
found on [GitHub](https://github.com/EvaMaeRey/ggcirclepack).

```{r contents_3A, dpi = 96, fig.width = 12, fig.height = 6, fig.alt = 'position_circlepack() and position_circlepackdodge()'}
# position_circlepack()
plot <- ggplot(dat, aes(x = x, y = y, size = size)) +
  geom_point(alpha = 0.25,
             position = position_circlepack(density = 0.25, 
                                            aspect_ratio = 1)) +
  coord_equal(xlim = c(-1, 1), 
              ylim = c(-1.1, 1.1)) +
  theme_bw() +
  theme(legend.position = "none",
        axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("position_circlepack()")
  
# position_circlepackdodge()
dodged_plot <- ggplot(dat, aes(x = x, y = y, color = group, size = size)) +
  geom_point(alpha = 0.25,
             position = position_circlepackdodge(width = 2, 
                                                 density = 0.25, 
                                                 aspect_ratio = 1)) +
  coord_equal(xlim = c(-1, 1), 
              ylim = c(-1.1, 1.1)) +
  theme_bw() + 
  theme(legend.position = "none",
        axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("position_circlepackdodge()")
  
plot + dodged_plot
```

Like `position_sunflower()`, `position_circlepack()` works from the inside out 
in the order of the data. So arranging the data by size organizes the points 
accordingly.

```{r contents_3B, dpi = 96, fig.width = 12, fig.height = 4, fig.alt = 'random, ascending, descending'}
# random size, base plot
random <- ggplot(dat, aes(x = x, y = y, size = size)) +
  geom_point(alpha = 0.25,
             position = position_circlepack(density = 0.075, 
                                            aspect_ratio = 1)) +
  coord_equal(xlim = c(-1, 1), 
              ylim = c(-1.1, 1.1)) +
  theme_bw() +
  theme(legend.position = "none",
        axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("random")

# ascending size
ascending <- random %+% 
  arrange(dat, size) + 
  ggtitle("ascending")

# descending size
descending <- random %+% 
  arrange(dat, desc(size)) + 
  ggtitle("descending")

random + ascending + descending
```

## Example

The current version of `vayr` also includes data from the Patriot Act experiment 
described in *Persuasion in Parallel*. The Patriot Act was an anti-terrorism law,
and the `patriot_act` dataset contains data relating to an experiment that measured 
support for this law after exposing participants to statements that cast it in either 
a negative or positive light. The experiment was conducted in 2009 with a nationwide 
sample, and it was replicated in 2015 with a sample of MTurkers. In both instances, 
the treatments had a parallel effect on Democrats and Republicans. There are four 
variables in the data: 

  + `sample_label`, the study to which the participant belonged
  + `pid_3`, the partisanship of the participant
  + `T1_content`, the statements to which the participant was exposed
  +  `PA_support`, the participant's post-treatment support for the Patriot Act 
  
The data is visualized below using `position_sunflowerdodge()` from `vayr`, 
demonstrating the package's functionality with a real-world example.

```{r patriot_act_visualization, dpi = 96, fig.width = 10, fig.height = 6, fig.alt = "patriot_act"}
# A df for statistical models
summary_df <- patriot_act |> 
  group_by(T1_content, pid_3, sample_label) |> 
  reframe(tidy(lm_robust(PA_support~1)))

# A df for direct labels
label_df <- summary_df |> 
  filter(sample_label == "Original Study", T1_content == "Control") |> 
  mutate(PA_support = case_when(pid_3 == "Democrat"~conf.low - 0.15, 
                                pid_3 == "Republican"~conf.high + 0.15))

ggplot(patriot_act, aes(T1_content, PA_support, color = pid_3, group = pid_3)) +
  # the data
  geom_point(position = position_sunflowerdodge(width = 0.5, density = 10),
             size = 0.1, alpha = 0.5) +
  # the statistical model
  geom_line(data = summary_df, aes(x = T1_content, y = estimate),  
            position = position_dodge(width = 0.5), linewidth = 0.5) +  
  geom_point(data = summary_df, aes(x = T1_content, y = estimate),  
             position = position_dodge(width = 0.5), size = 3) +
  geom_linerange(data = summary_df, aes(x = T1_content, y = estimate,
                                        ymin = conf.low, ymax = conf.high),
                 position = position_dodge(width = 0.5)) +
  # the direct labels
  geom_text(data = label_df, aes(label = pid_3)) +
  # the rest
  scale_color_manual(values = c("blue4", "red3")) +
  scale_y_continuous(breaks = 1:7) +
  coord_equal() +
  facet_wrap(~sample_label) +
  theme_bw() +
  theme(legend.position = "none",
        strip.background = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(y = "Do you oppose or support the Patriot Act?
            [1: Oppose very strongly to 7: Support very strongly]",
       x = "Randomly assigned information")
```
