---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# vayr

<!-- badges: start -->
[![R-CMD-check](https://github.com/acoppock/vayr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/acoppock/vayr/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/acoppock/vayr/graph/badge.svg)](https://app.codecov.io/gh/acoppock/vayr)
[![CRAN status](https://www.r-pkg.org/badges/version/vayr)](https://CRAN.R-project.org/package=vayr)
[![Downloads](https://cranlogs.r-pkg.org/badges/grand-total/vayr)](https://CRAN.R-project.org/package=vayr)
<!-- badges: end -->

The goal of `vayr` is to provide `ggplot2` extensions that foster "visualize as
you randomize" principles. These principles are outlined in detail in "Visualize
As You Randomize: Design-based Statistical Graphs for Randomized Experiments,"
a chapter in *Advances in Experimental Political Science*
([PDF](https://alexandercoppock.com/coppock_2021.pdf),
[DOI](https://doi.org/10.1017/9781108777919.022)). The package includes position
adjustments that avoid over-plotting, which helps organize "data-space" to better
contextualize statistical models.

## Installation

The release version of `vayr` can be installed from CRAN, and the development
version can be installed from [GitHub](https://github.com/) using a package like
`remotes`, `devtools`, or `pak`. `vayr` relies on `ggplot2`, `packcircles`, and
`withr`, so these must be installed as well.

``` r
# From CRAN
install.packages("vayr")

# From GitHub
# install.packages("pak")
pak::pak("acoppock/vayr")
```

## Position adjustments

`vayr` provides six position adjustments that apply to "point-like" geoms such as
`geom_point()` and `geom_text()`. They come in pairs, one that arranges
over-plotted points and one that also dodges groups side-to-side:

  + `position_jitter_ellipse()` and `position_jitterdodge_ellipse()` sample from
    an elliptical field rather than the rectangle that `position_jitter()` uses,
    so the dispersion retains the impression of a single point.
  + `position_sunflower()` and `position_sunflowerdodge()` arrange over-plotted
    points in a sunflower pattern, working from the inside out in the order of
    the data. A point with nothing over-plotting it stays where it is.
  + `position_circlepack()` and `position_circlepackdodge()` pack over-plotted
    points of varying sizes into an elliptical area, which is useful when point
    size carries a weight.

Each takes a `density` argument controlling how tightly the points pack, and an
`aspect_ratio` argument that compensates for a non-square plotting region.

```{r overview, dpi = 96, fig.width = 12, fig.height = 8, fig.alt = 'the six position adjustments applied to 200 points over-plotted at the origin'}
library(ggplot2)
library(patchwork)
library(vayr)

set.seed(1)

dat <- data.frame(
  x = rep(0, 200),
  y = rep(0, 200),
  group = rep(c("A", "B", "B", "B"), 50),
  size = runif(200, 0, 1)
)

vayr_theme <- list(
  theme_bw(),
  theme(legend.position = "none",
        axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 11))
)

jitter_ellipse <- ggplot(dat, aes(x, y)) +
  geom_point(position = position_jitter_ellipse(width = 0.5, height = 0.5)) +
  coord_equal(xlim = c(-1.1, 1.1), ylim = c(-1.1, 1.1)) +
  vayr_theme + ggtitle("position_jitter_ellipse()")

jitterdodge_ellipse <- ggplot(dat, aes(x, y, color = group)) +
  geom_point(position = position_jitterdodge_ellipse(dodge.width = 2,
                                                     jitter.width = 0.5,
                                                     jitter.height = 0.5)) +
  coord_equal(xlim = c(-1.1, 1.1), ylim = c(-1.1, 1.1)) +
  vayr_theme + ggtitle("position_jitterdodge_ellipse()")

sunflower <- ggplot(dat, aes(x, y)) +
  geom_point(position = position_sunflower(density = 1, aspect_ratio = 1)) +
  coord_equal(xlim = c(-2.1, 2.1), ylim = c(-2.1, 2.1)) +
  vayr_theme + ggtitle("position_sunflower()")

sunflowerdodge <- ggplot(dat, aes(x, y, color = group)) +
  geom_point(position = position_sunflowerdodge(width = 4, density = 1,
                                                aspect_ratio = 1)) +
  coord_equal(xlim = c(-2.1, 2.1), ylim = c(-2.1, 2.1)) +
  vayr_theme + ggtitle("position_sunflowerdodge()")

circlepack <- ggplot(dat, aes(x, y, size = size)) +
  geom_point(alpha = 0.25,
             position = position_circlepack(density = 0.25, aspect_ratio = 1)) +
  coord_equal(xlim = c(-1.1, 1.1), ylim = c(-1.1, 1.1)) +
  vayr_theme + ggtitle("position_circlepack()")

circlepackdodge <- ggplot(dat, aes(x, y, color = group, size = size)) +
  geom_point(alpha = 0.25,
             position = position_circlepackdodge(width = 2, density = 0.25,
                                                 aspect_ratio = 1)) +
  coord_equal(xlim = c(-1.1, 1.1), ylim = c(-1.1, 1.1)) +
  vayr_theme + ggtitle("position_circlepackdodge()")

(jitter_ellipse + sunflower + circlepack) /
  (jitterdodge_ellipse + sunflowerdodge + circlepackdodge)
```

## Learn more

The vignette walks through every position adjustment, explains how `density` and
`aspect_ratio` interact with the plotting region, and closes with a worked
example that plots an experiment's data and its statistical model together:

``` r
vignette("vayr-vignette", package = "vayr")
```

It is also on the package site, along with the reference documentation:
<https://alexandercoppock.com/vayr/>.

## Citation

``` r
citation("vayr")
```

Coppock, Alexander. 2021. "Visualize As You Randomize: Design-based Statistical
Graphs for Randomized Experiments." In *Advances in Experimental Political
Science*, edited by James N. Druckman and Donald P. Green, 320–336. New York:
Cambridge University Press. <https://doi.org/10.1017/9781108777919.022>
