---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# vayr

<!-- badges: start -->
<!-- badges: end -->

The goal of `vayr` is to provide `ggplot2` extensions that foster "visualize as 
you randomize" principles. These principles are outlined in detail in "Visualize 
as You Randomize: Design-Based Statistical Graphs for Randomized Experiments." 
That paper can be accessed as a [PDF](https://alexandercoppock.com/coppock_2020.pdf).
The package includes position adjustments to avoid over-plotting, facilitating 
the the organization of "data-space" to better contextualize statistical models.

## Installation

The release version of `vayr` can be installed from CRAN, and the development 
version can be installed from [GitHub](https://github.com/) using a package like 
`remotes`, `devtools `, or `pak`. It is important to note that `vayr` relies on 
`ggplot2`, `packcircles`, and `withr`, so these must be installed as well.

``` r
# From CRAN
install.packages("vayr")

# From GitHub
# install.packages("pak")
pak::pak("acoppock/vayr")
```

## Contents

`vayr` contains a handful of `ggplot2` functions that apply as position adjustments 
to "point-like" geoms such as `geom_point` or `geom_text`:

  + `position_jitter_ellipse()` and `position_jitterdodge_ellipse()`
  + `position_sunflower()` and `position_sunflowerdodge()`
  + `position_circlepack()` and `position_circlepackdodge()`
  
These functions avoid over-plotting, so they are especially useful when plotting 
discrete rather than continuous data. As a matter of demonstration, they are used 
below to visualize synthetic data, over-plotted at the origin.

```{r, echo = FALSE}
suppressMessages(library(dplyr))
```

```{r setup}
library(dplyr)
library(estimatr)
library(ggplot2)
library(patchwork)
library(vayr)

dat <- data.frame(
  x = c(rep(0, 200)),
  y = c(rep(0, 200)),
  group = (rep(c("A", "B", "B", "B"), 50)),
  size = runif(200, 0, 1)
)
```

If position is the product of discrete variables alone, then over-plotting is of 
particular concern. `position_jitter()` can mitigate this concern. It introduces 
variation by randomly sampling points on a rectangle. This approach is effective 
but can be unattractive. The position adjustments in `vayr` are attempts to do better. 

```{r contents_0, dpi = 96, fig.width = 12, fig.height = 6, fig.alt = 'perfect over-plotting and position_jitter()'}
# perfectly over-plotted points
over_plot <- ggplot(dat, aes(x = x, y = y)) +
  geom_point() +
  coord_equal(xlim = c(-1.1, 1.1), 
              ylim = c(-1.1, 1.1)) +
  theme_bw() +
  theme(axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle('"perfect over-plotting"')

# position_jitter()
jitter_plot <- ggplot(dat, aes(x = x, y = y)) + 
  geom_point(position = position_jitter(width = 0.5, 
                                        height = 0.5)) +
  coord_equal(xlim = c(-1.1, 1.1), 
              ylim = c(-1.1, 1.1)) +
  theme_bw() +
  theme(axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("position_jitter()")

over_plot + jitter_plot
```

### Position Jitter Ellipse

This position adjustment adds elliptical random noise to perfectly over-plotted 
points, offering a pleasing way to visualize many points that represent the same 
position. The benefit of sampling on an ellipse of a given `height` and `width` 
rather than on a rectangle is that the resulting dispersion retains the impression 
of a single point. The size of the ellipses stays constant, while their density 
varies depending on the amount of data.

```{r contents_1, dpi = 96, fig.width = 12, fig.height = 6, fig.alt = 'position_jitter_ellipse() and position_jitterdodge_ellipse()'}
# position_jitter_ellipse()
jitter_ellipse_plot <- ggplot(dat, aes(x = x, y = y)) +
  geom_point(position = position_jitter_ellipse(width = 0.5, 
                                                height = 0.5)) +
  coord_equal(xlim = c(-1.1, 1.1), 
              ylim = c(-1.1, 1.1)) +
  theme_bw() +
  theme(axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("position_jitter_ellipse()")

# position_jitterdodge_ellipse()
jitterdodge_ellipse_plot <- ggplot(dat, aes(x = x, y = y, color = group)) +
  geom_point(position = position_jitterdodge_ellipse(dodge.width = 2, 
                                                     jitter.width = 0.5, 
                                                     jitter.height = 0.5)) +
  coord_equal(xlim = c(-1.1, 1.1), 
              ylim = c(-1.1, 1.1)) +
  theme_bw() +
  theme(legend.position = "none",
        axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("position_jitterdodge_ellipse()")
  
jitter_ellipse_plot + jitterdodge_ellipse_plot
```

### Position Sunflower

This position adjustment arranges perfectly over-plotted points using a sunflower 
algorithm, which produces a pattern that resembles the seeds of a sunflower, 
working from the inside out in the order of the data. The parameters for this 
position adjustment are `density` and `aspect_ratio`. The size of the flowers 
varies depending on the amount over over-plotting, but the density of the pattern 
remains constant. It is generally recommended that the position adjustment be 
used along with `coord_equal()`, in which case the default aspect ratio of 1 yields 
perfectly circular flowers, but the aspect ratio of the flowers can be adjusted
if need be.

```{r contents_2A, dpi = 96, fig.width = 12, fig.height = 6, fig.alt = 'position_sunflower() and position_sunflowerdodge()'}
# position_sunflower()
sunflower_plot <- ggplot(dat, aes(x = x, y = y)) +
  geom_point(position = position_sunflower(density = 1, 
                                           aspect_ratio = 1)) +
  coord_equal(xlim = c(-2.1, 2.1), 
              ylim = c(-2.1, 2.1)) +
  theme_bw() +
  theme(axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("position_sunflower()")
  
# position_sunflowerdodge()
sunflowerdodge_plot <- ggplot(dat, aes(x = x, y = y, color = group)) +
  geom_point(position = position_sunflowerdodge(width = 4, 
                                                density = 1, 
                                                aspect_ratio = 1)) +
  coord_equal(xlim = c(-2.1, 2.1), 
              ylim = c(-2.1, 2.1)) +
  theme_bw() + 
  theme(legend.position = "none",
        axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("position_sunflowerdodge()")
  
sunflower_plot + sunflowerdodge_plot
```

The `density` parameter controls the density of the pattern. A density of 1 is 
normalized to 100 points in a unit circle; a density of 2, 200 points; and a 
density of 0.5, 50 points. Because density is normalized relative to Cartesian 
units, its visual effect depends on the ranges of the axes and the dimensions of 
the saved image. Smaller ranges or larger dimensions require a greater density to 
produce the same visual effect. The size of points should also be taken into account.

``` {r contents_2B, dpi = 96, fig.width = 12, fig.height = 12, fig.alt = 'density', echo = FALSE}
densities <- rep(c(0.5, 1, 2), each = 3)
ns <- rep(c(50, 100, 200), 3)

density_plots <- list()

for (i in 1:9) {
  density_dat <- data.frame(x = c(rep(0, ns[i])), y = c(rep(0, ns[i])))

  density_plots[[i]] <- ggplot(density_dat, aes(x, y)) +
    geom_point(position = position_sunflower(density = densities[i])) +
    coord_equal(xlim = c(-2, 2),
                ylim = c(-2, 2)) +
    theme_bw() +
    labs(title = paste0("n = ", ns[i], ", density = ", densities[i])) +
    theme(axis.title = element_blank(),
          panel.grid.minor = element_blank(),
          plot.title = element_text(face = "bold"))
}

wrap_plots(density_plots, ncol = 3)
```

The `aspect_ratio` parameter changes the aspect ratio of the flowers, which is
their width divided by their height. This is useful when using the position 
adjustment without `coord_equal()`. The flowers can be made wider or taller to 
compensate for the aspect ratio of the axes or the image. The aspect ratio of the 
flowers should be set to the opposite of the aspect ratio for which it must compensate. 
For instance, consider a plot with anx axis that ranges from 0 to 1, and a y axis 
that ranges from 0 to 2. Saving this plot as a square image would  squish the y 
axis, resulting in wider flowers. Setting the aspect ratio of the flowers to 0.5 
would offset this distortion.

``` {r contents_2C, dpi = 96, fig.width = 12, fig.height = 12, fig.alt = 'aspect_ratio', echo = FALSE}
flower_ratios <- rep(c(0.5, 1, 2), each = 3)
axis_ratios <- rep(c(2, 1, 0.5), 3)

aspect_ratio_dat <- data.frame(x = c(rep(0, 100)), y = c(rep(0, 100)))

aspect_ratio_plots <- list()

for (i in 1:9) {

  aspect_ratio_plots[[i]] <- ggplot(dat, aes(x, y)) +
    geom_point(position = position_sunflower(aspect_ratio = flower_ratios[i])) +
    coord_fixed(xlim = c(-2, 2),
                ylim = c(-2, 2),
                ratio = axis_ratios[i]) +
    theme_bw() +
    labs(title = paste0("aspect_ratio = ", flower_ratios[i])) +
    theme(axis.title = element_blank(),
          panel.grid.minor = element_blank(),
          plot.title = element_text(face = "bold"))
}

wrap_plots(aspect_ratio_plots, ncol = 3)
```

### Position Circle Pack

This position adjustment uses circle packing algorithms from  `packcircles` to 
arrange perfectly over-plotted points of varying sizes into an elliptical area. 
It also takes `density` and `aspect_ratio` as parameters. This position adjustment 
should not be confused with `geom_circlepack()` from `ggcirclepack` which can be 
found on [GitHub](https://github.com/EvaMaeRey/ggcirclepack).

```{r contents_3A, dpi = 96, fig.width = 12, fig.height = 6, fig.alt = 'position_circlepack() and position_circlepackdodge()'}
# position_circlepack()
circlepack_plot <- ggplot(dat, aes(x = x, y = y, size = size)) +
  geom_point(alpha = 0.25,
             position = position_circlepack(density = 0.25, 
                                            aspect_ratio = 1)) +
  coord_equal(xlim = c(-1, 1), 
              ylim = c(-1.1, 1.1)) +
  theme_bw() +
  theme(legend.position = "none",
        axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("position_circlepack()")
  
# position_circlepackdodge()
circlepackdodge_plot <- ggplot(dat, aes(x = x, y = y, color = group, size = size)) +
  geom_point(alpha = 0.25,
             position = position_circlepackdodge(width = 2, 
                                                 density = 0.25, 
                                                 aspect_ratio = 1)) +
  coord_equal(xlim = c(-1, 1), 
              ylim = c(-1.1, 1.1)) +
  theme_bw() + 
  theme(legend.position = "none",
        axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("position_circlepackdodge()")
  
circlepack_plot + circlepackdodge_plot
```

Like `position_sunflower()` `position_circlepack()` works from the inside out 
in the order of the data. So arranging the data by size organizes the points 
accordingly.

```{r contents_3B, dpi = 96, fig.width = 12, fig.height = 4, fig.alt = 'random, ascending, descending'}
# random size, base plot
random <- ggplot(dat, aes(x = x, y = y, size = size)) +
  geom_point(alpha = 0.25,
             position = position_circlepack(density = 0.075, 
                                            aspect_ratio = 1)) +
  coord_equal(xlim = c(-1, 1), 
              ylim = c(-1.1, 1.1)) +
  theme_bw() +
  theme(legend.position = "none",
        axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("random")

# ascending size
ascending <- random %+% 
  arrange(dat, size) + 
  ggtitle("ascending")

# descending size
descending <- random %+% 
  arrange(dat, desc(size)) + 
  ggtitle("descending")

random + ascending + descending
```

## Example

`vayr` also includes data from the Patriot Act experiment described in 
[*Persuasion in Parallel*](https://alexandercoppock.com/coppock_2022.html).
The Patriot Act was an anti-terrorism law, and the `patriot_act` dataset contains 
data relating to an experiment that measured support for this law after randomly 
exposing participants to statements that cast the legislation in either a negative 
or positive light. The experiment was conducted in 2009 with a nationwide sample, 
and it was replicated in 2015 with a sample of MTurkers. In both instances, the 
treatments had a similar effect on Democrats and Republicans. There are four 
variables in the data: 

  + `sample_label`, the study to which the participant belonged
  + `pid_3`, the partisanship of the participant
  + `T1_content`, the statements to which the participant was exposed
  +  `PA_support`, the participant's post-treatment support for the Patriot Act 
  
The data is visualized below using `position_sunflowerdodge()` from `vayr`. Note
that both `density` and `aspect_ratio` are adjusted. A high `density` is needed
because of a small point size, and a tall `aspect_ratio` accounts for a wide plot.

```{r patriot_act_visualization, dpi = 96, fig.width = 12, fig.height = 6, fig.alt = "patriot_act"}
# A df for statistical models
summary_df <- patriot_act |>
  group_by(T1_content, pid_3, sample_label) |>
  reframe(tidy(lm_robust(PA_support ~ 1)))

# A df for direct labels
label_df <- summary_df |>
  filter(sample_label == "Original Study", T1_content == "Control") |>
  mutate(
    PA_support = case_when(
      pid_3 == "Democrat" ~ conf.low - 0.15,
      pid_3 == "Republican" ~ conf.high + 0.15
    )
  )

ggplot(patriot_act, aes(T1_content, PA_support, color = pid_3, group = pid_3)) +
  # the data
  geom_point(position = position_sunflowerdodge(width = 0.5, 
                                                density = 50,
                                                aspect_ratio = 0.5),
             size = 0.1, alpha = 0.5) +
  # the statistical model
  geom_line(data = summary_df, aes(x = T1_content, y = estimate),  
            position = position_dodge(width = 0.5), linewidth = 0.5) +  
  geom_point(data = summary_df, aes(x = T1_content, y = estimate),  
             position = position_dodge(width = 0.5), size = 3) +
  geom_linerange(data = summary_df, aes(x = T1_content, y = estimate,
                                        ymin = conf.low, ymax = conf.high),
                 position = position_dodge(width = 0.5)) +
  # the direct labels
  geom_text(data = label_df, aes(label = pid_3)) +
  # the rest
  scale_color_manual(values = c("blue4", "red3")) +
  scale_y_continuous(breaks = 1:7) +
  coord_fixed(ratio = 0.5) + # ratio for coord_fixed is y/x rather than x/y
  facet_wrap(~sample_label) +
  theme_bw() +
  theme(legend.position = "none",
        strip.background = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(y = "Do you oppose or support the Patriot Act?
            [1: Oppose very strongly to 7: Support very strongly]",
       x = "Randomly assigned information")
```
