---
title: "Using vayr with ggplot2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vayr-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

# vayr

The goal of 'vayr' is to provide 'ggplot2' extensions that foster "visualize as you randomize" principles,
which guide the visualization of experimental data. Thus far, the package includes position adjustments
to avoid over-plotting, facilitating the plotting of statistical models in "data-space." In other words,
these position adjustments help you organize "data-space," such that it can contextualize your models.
The principles underlying 'vayr' are outlined in detail in 
"Visualize as You Randomize: Design-Based Statistical Graphs for Randomized Experiments."
That paper can be accessed as a [PDF](https://alexandercoppock.com/coppock_2020.pdf).


## Installation

At the moment, 'vayr' only exists on [GitHub](https://github.com/), 
although we ultimately hope to release the package to CRAN.
You can install the development version of the package using 'pak,'
as shown below. It is important to note that 'vayr' relies on
'ggplot2,' 'packcircles,' and 'withr,' so these packages must also be installed.

``` r
# install.packages("pak")
pak::pak("acoppock/vayr")
```

## Contents

The current version of 'vayr' contains a handful new 'ggplot2' functions, all of which may be applied 
as position adjustments to the point geom: **position_jitter_ellipse()**,
**position_sunflower()**, and **position_circlepack()**; as well as their dodged counterparts: 
**position_jitterdodge_ellipse()**, **position_sunflowerdodge()**, and **position_circlepackdodge()**. 
These functions avoid perfect over-plotting and are therefore useful when plotting discrete rather than continuous data
in a bivariate context. As a matter of demonstration, they are used below to visualize synthetic data,
over-plotted at the origin.

```{r, echo = FALSE}
suppressMessages(library(dplyr))
```

```{r setup}
library(dplyr)
library(ggplot2)
library(patchwork)
library(vayr)
library(estimatr)

dat <- data.frame(
  X = c(rep(0, 200)),
  Y = c(rep(0, 200)),
  Group = (rep(c("A", "B", "B", "B"), 50)),
  Size = runif(200, 0, 1),
  Enum = 1:200
)
```

If position is the product of discrete variables alone, then perfect over-plotting is of particular concern. 
To mitigate this concern, **position_jitter()** can introduce a degree of variation, 
whereby points are randomly sampled on a rectangle, but it dilutes the visual effect of 
discrete data. The functions in 'vayr' are preferable alternatives.

```{r contents_0, dpi = 300, fig.width = 12, fig.height = 6, fig.alt = 'perfect over-plotting and position_jitter()'}
# perfectly over-plotted points
overplot <- ggplot(dat, aes(x = X, y = Y)) +
  geom_point() +
  coord_equal(xlim = c(-1.1, 1.1), ylim = c(-1.1, 1.1)) +
  theme_bw() +
  theme(axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle('"perfect over-plotting"')

# position_jitter()
plot <- ggplot(dat, aes(x = X, y = Y)) + 
  geom_point(position = position_jitter(width = 0.5, height = 0.5)) +
  coord_equal(xlim = c(-1.1, 1.1), ylim = c(-1.1, 1.1)) +
  theme_bw() +
  theme(axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("position_jitter()")


overplot + plot
```

### Position Jitter Ellipse

This position adjustment adds elliptical random noise to perfectly over-plotted points, offering
a pleasing way to visualize many points that represent the same position. The benefit of
sampling on an ellipse (of a given **height** and **width**) rather than a rectangle is that the 
resulting dispersion retains the impression of a single point. The size of these meta-points is constant across a graph,
while the density varies depending on the amount of over-plotting.

```{r contents_1, dpi = 300, fig.width = 12, fig.height = 6, fig.alt = 'position_jitter_ellipse() and position_jitterdodge_ellipse()'}
# position_jitter_ellipse()
plot <- ggplot(dat, aes(x = X, y = Y)) +
  geom_point(position = position_jitter_ellipse(width = 0.5, height = 0.5)) +
  coord_equal(xlim = c(-1.1, 1.1), ylim = c(-1.1, 1.1)) +
  theme_bw() +
  theme(axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("position_jitter_ellipse()")

# position_jitterdodge_ellipse()
dodged_plot <- ggplot(dat, aes(x = X, y = Y, color = Group)) +
  geom_point(position = position_jitterdodge_ellipse(dodge.width = 2, 
                                                     jitter.width = 0.5, 
                                                     jitter.height = 0.5)) +
  coord_equal(xlim = c(-1.1, 1.1), ylim = c(-1.1, 1.1)) +
  theme_bw() +
  theme(legend.position = "none",
        axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("position_jitterdodge_ellipse()")
  
plot + dodged_plot
```

### Position Sunflower

This position adjustment arranges perfectly over-plotted points using a sunflower algorithm, 
which produces a pattern that resembles the seeds of a sunflower. The algorithm works 
from the inside out in the order of the data. The parameters for this position adjustment 
(**density** and **aspect_ratio**) are novel but intuitive. The size of the meta-points 
(the flowers, that is) varies depending on the amount over over-plotting, but the density of
the pattern remains constant. It is possible to adjust the aspect ratio of the meta-points,
but it is generally recommended that the position adjustment be used in tandem with **coord_equal()**, 
in which case an aspect ratio of 1 (the default) should yield perfect circles.


```{r contents_2A, dpi = 300, fig.width = 12, fig.height = 6, fig.alt = 'position_sunflower() and position_sunflowerdodge()'}
# position_sunflower()
plot <- ggplot(dat, aes(x = X, y = Y)) +
  geom_point(position = position_sunflower(density = 1, aspect_ratio = 1)) +
  coord_equal(xlim = c(-1.1, 1.1), ylim = c(-1.1, 1.1)) +
  theme_bw() +
  theme(axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("position_sunflower()")
  
# position_sunflowerdodge()
dodged_plot <- ggplot(dat, aes(x = X, y = Y, color = Group)) +
  geom_point(position = position_sunflowerdodge(width = 2, density = 1, aspect_ratio = 1)) +
  coord_equal(xlim = c(-1.1, 1.1), ylim = c(-1.1, 1.1)) +
  theme_bw() + 
  theme(legend.position = "none",
        axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("position_sunflowerdodge()")
  
plot + dodged_plot
```

As stated, the density of the seeds can be adjusted. Density is not standardized
but rather depends on the ranges of the axes and dimensions of the image. Loosely,
it can be thought of as the number of points within a certain Cartesian area.
Thus smaller ranges and/or larger dimensions will require a greater density to produce the same
visual effect.

```{r contents_2B, dpi = 300, fig.width = 12, fig.height = 4, fig.alt = 'density'}
# base plot
plot <- ggplot(dat, aes(x = X, y = Y)) +
  coord_equal(xlim = c(-1.1, 1.1), ylim = c(-1.1, 1.1)) +
  theme_bw() +
  theme(axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold"))

# low density
low <- plot + geom_point(position = position_sunflower(density = 0.5, aspect_ratio = 1),
                         size = 0.25) + ggtitle("density = 0.5")

# medium density
medium <- plot + geom_point(position = position_sunflower(density = 1, aspect_ratio = 1),
                            size = 0.25) + ggtitle("density = 1")

# high density
high <- plot + geom_point(position = position_sunflower(density = 2, aspect_ratio = 1),
                          size = 0.25) + ggtitle("density = 2")

low + medium + high
```

The aspect ratio of the flowers can also be tweaked. This is useful when the position adjustment
is used without **coord_equal()**. The flowers can be made wider or taller to compensate for the
aspect ratio of the axes and/or the image, in which case the aspect ratio of the flower should be 
set to the opposite of the aspect ratio for which it must compensate. For instance, saving a plot with a 
vertical range of 2 and a horizontal range of 1 as a square image would result in wide flowers. 
Setting the aspect ratio of the flowers to 1/2 would offset the distortion, although refinement 
would likely be warranted due to labels, legends, and the like.

```{r contents_2C, dpi = 300, fig.width = 12, fig.height = 4, fig.alt = 'aspect_ratio'}
# base plot
plot <- ggplot(dat, aes(x = X, y = Y)) +
  coord_equal(xlim = c(-1.1, 1.1), ylim = c(-1.1, 1.1)) +
  theme_bw() +
  theme(axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold"))

# wide aspect ratio 
wide <- plot + geom_point(position = position_sunflower(density = 1, aspect_ratio = 0.5),
                         size = 0.25) + ggtitle("aspect_ratio = 0.5")

# normal aspect ratio
normal <- plot + geom_point(position = position_sunflower(density = 1, aspect_ratio = 1),
                            size = 0.25) + ggtitle("aspect_ratio = 1")

# tall aspect ratio
tall <- plot + geom_point(position = position_sunflower(density = 1, aspect_ratio = 2),
                          size = 0.25) + ggtitle("aspect_ratio = 2")

wide + normal + tall
```

### Position Circle Pack

This position adjustment uses circle packing algorithms from the 'packcircles' package
to arrange perfectly over-plotted points of varying sizes into an elliptical area. 
It also takes **density** and **aspect_ratio** as parameters, and they function as
previously described. Again, the desirable density will depend on both the ranges of the axes 
and the dimensions of the image.

```{r contents_3A, dpi = 300, fig.width = 12, fig.height = 6, fig.alt = 'position_circlepack() and position_circlepackdodge()'}
# position_circlepack()
plot <- ggplot(dat, aes(x = X, y = Y, size = Size)) +
  geom_point(position = position_circlepack(density = 0.25, aspect_ratio = 1),
             alpha = 0.25) +
  coord_equal(xlim = c(-1, 1), ylim = c(-1.1, 1.1)) +
  theme_bw() +
  theme(legend.position = "none",
        axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("position_circlepack()")
  
# position_circlepackdodge()
dodged_plot <- ggplot(dat, aes(x = X, y = Y, color = Group, size = Size)) +
  geom_point(position = position_circlepackdodge(width = 2, density = 0.25, aspect_ratio = 1),
             alpha = 0.25) +
  coord_equal(xlim = c(-1, 1), ylim = c(-1.1, 1.1)) +
  theme_bw() + 
  theme(legend.position = "none",
        axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("position_circlepackdodge()")
  
plot + dodged_plot
```

Like **position_sunflower()**, **position_circlepack()** creates meta-points from
the inside out in the order of the data. Thus, arranging the data by size will
organize the meta-points accordingly.

```{r contents_3B, dpi = 300, fig.width = 12, fig.height = 4, fig.alt = 'random, ascending, descending'}
# random size, base plot
random <- dat |> ggplot(aes(x = X, y = Y, size = Size)) +
  geom_point(position = position_circlepack(density = 0.075, aspect_ratio = 1),
             alpha = 0.25) +
  coord_equal(xlim = c(-1, 1), ylim = c(-1.1, 1.1)) +
  theme_bw() +
  theme(legend.position = "none",
        axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("random")

# ascending size
ascending <- random %+% (dat |> arrange(Size)) + ggtitle("ascending")

# descending size
descending <- random %+% (dat |> arrange(desc(Size))) + ggtitle("descending")

random + ascending + descending
```

## Example

The current version of 'vayr' also includes data from the Patriot Act experiment described in
*Persuasion in Parallel*. The Patriot Act was an anti-terrorism law; the eponymous dataset,
**patriot_act**, contains data relating to an experiment that measured support for this law after
exposing participants to statements that cast it in either a negative or positive light. 
The experiment was conducted in 2009 with a nationwide sample, and it was replicated in 2015 with a 
sample of MTurkers. In both instances, the treatments had a parallel effect on Democrats and Republicans.
There are four variables in the data: 
**sample_label**, the study to which the participant belonged (Original Study or Mechanical Turk Replication); 
**pid_3**, the partisanship of the participant (Republican or Democrat); 
**T1_content**, the statements to which the participant was exposed (Control, Pro, or Con); 
and **PA_support**, the participant's post-treatment support for the Patriot Act (1: Oppose very strongly to 7: Support very strongly).
The data is visualized below using **position_sunflowerdodge()** from 'vayr' 
so as to demonstrate the package's functionality with a real-world example.

```{r patriot_act_visualization, dpi = 300, fig.width = 10, fig.height = 6, fig.alt = "patriot_act"}
# A df for statistical models
summary_df <- patriot_act |> 
  group_by(T1_content, pid_3, sample_label) |> 
  reframe(tidy(lm_robust(PA_support~1)))

# A df for direct labels
label_df <- summary_df |> 
  filter(sample_label == "Original Study", T1_content == "Control") |> 
  mutate(PA_support = case_when(pid_3 == "Democrat"~conf.low - 0.15, 
                                pid_3 == "Republican"~conf.high + 0.15))

ggplot(patriot_act, aes(T1_content, PA_support, color = pid_3, group = pid_3)) +
  # the data
  geom_point(position = position_sunflowerdodge(width = 0.5, density = 10),
             size = 0.1, alpha = 0.5) +
  # the statistical model
  geom_line(data = summary_df, aes(x = T1_content, y = estimate),  
            position = position_dodge(width = 0.5), linewidth = 0.5) +  
  geom_point(data = summary_df, aes(x = T1_content, y = estimate),  
             position = position_dodge(width = 0.5), size = 3) +
  geom_linerange(data = summary_df, aes(x = T1_content, ymin = conf.low, ymax = conf.high, y = estimate),
                 position = position_dodge(width = 0.5)) +
  # the direct labels
  geom_text(data = label_df, aes(label = pid_3)) +
  # the rest
  scale_color_manual(values = c("blue4", "red3")) +
  scale_y_continuous(breaks = 1:7) +
  coord_equal() +
  facet_wrap(~sample_label) +
  theme_bw() +
  theme(legend.position = "none",
        strip.background = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(y = "Do you oppose or support the Patriot Act?
            [1: Oppose very strongly to 7: Support very strongly]",
       x = "Randomly assigned information")
```
